NodIO: a framework for pool-based evolutionary computation

### 2016-03-16 EvoSoft16/nodio-evosoft16.tex
Starting the new paper for EvoSoft

\title{Performance for the masses: Experiments with A Web Based Architecture to Harness Volunteer Resources 
for Low Cost Distributed Evolutionary Computation}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{5} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
% \author{Juan-J.~Merelo, Dept. of Computer Architecture and Technology
%   and CITIC, University of Granada, Granada, Spain \\
% Mario Garc\'ia-Valdez, Dept. of Graduate Studies, Instituto
% Tecnológico de Tijuana, Tijuana, M\'exico\\
% Pedro A. Castillo, Dept. of Computer Architecture and Technology, University of Granada, Granada, Spain\\
% Pablo Garc\'ia-S\'anchez, Dept. of Computer Architecture and Technology, University of Granada, Granada, Spain\\
% Paloma de las Cuevas, Dept. of Computer Architecture and Technology,
% University of Granada, Granada, Spain\\

% \thanks{Corresponding author: Juan-J.~Merelo (Email: jmerelo@ugr.es).}%
% }
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{1932 Wallamaloo Lane}\\
       \affaddr{Wallamaloo, New Zealand}\\
       \email{trovato@corporation.com}
% 2nd. author
\alignauthor
G.K.M. Tobin\titlenote{The secretary disavows
any knowledge of this author's actions.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{P.O. Box 1212}\\
       \affaddr{Dublin, Ohio 43017-6221}\\
       \email{webmaster@marysville-ohio.com}
% 3rd. author
\alignauthor Lars Th{\o}rv{\"a}ld\titlenote{This author is the
one who did all the really hard work.}\\
       \affaddr{The Th{\o}rv{\"a}ld Group}\\
       \affaddr{1 Th{\o}rv{\"a}ld Circle}\\
       \affaddr{Hekla, Iceland}\\
       \email{larst@affiliation.org}
\and  % use '\and' if you need 'another row' of author names
% 4th. author
\alignauthor Lawrence P. Leipuner\\
       \affaddr{Brookhaven Laboratories}\\
       \affaddr{Brookhaven National Lab}\\
       \affaddr{P.O. Box 5000}\\
       \email{lleipuner@researchlabs.org}
% 5th. author
\alignauthor Sean Fogarty\\
       \affaddr{NASA Ames Research Center}\\
       \affaddr{Moffett Field}\\
       \affaddr{California 94035}\\
       \email{fogartys@amesres.org}
}

@@

Using volunteer's browsers as a computing resource presents several
advantages, but it remains a challenge to fully harness the browser's
capabilities and to model the user's behavior so that those
capabilities can be leveraged optimally. These are the objectives of
this paper, where we present the results of several evolutionary
computation experiments with different implementations of a volunteer
computing framework called {\sf NodIO}, designed to be easily
deployable on freely available cloud resources. We use different
implementations to find out which one is able to get the user to lend
more computing cycles and test different problems to check the
influence it has on said performance, as measured by the time needed
to find a solution, but also by the number of users engaged.  From
these experiments we can already draw some conclusions, besides the
fact that volunteer computing can be a valuable computing resource and
that it is essential to be as open as possible with software and data:
the user has to be kept engaged to obtain as many computing cycles as
possible, the client has to be built to use the computer capabilities
fully, and, finally, that the user contributions follow a common
statistical distribution.

@@

Nowadays there is a big and increasing number of distributed
evolutionary computation (EC) software frameworks available, usually
as open source frameworks \cite{Parejo12Survey}.  Most of them have
two features: they are {\em desktop} applications, that is, they must
be compiled or run directly on the operating system layer and, in addition,
they assume that the nodes that are going to run the system are under
the control of developers or they at least have access to them.

There is, then, space for other EC frameworks that do not make any of
these assumptions, such as {\sf NodIO} presented in
\cite{2016arXiv160101607Manom}. {\sf NodIO} implements a client-server application; the server offers the possibility to program an
evolutionary algorithm that is embedded in the served pages, and those pages can communicate between each other via the server. On the other hand, the client part can run in any
web-enabled device, including many over which developers have no
control, that is, volunteers.

We were interested in this kind of systems because, during the big
global crisis, our group faced the problem of diminishing funds for
buying new hardware. This was aggravated by the increasing maintenance costs and
extended downtime resulting from the continuous failures of existing
clusters.
Considering this, we leveraged our experience in the design of web
applications since their inception, including JavaScript, and other
volunteer and
unconventional distributed evolutionary computing systems to 
design and release a new free framework that would allow anyone to
create a volunteer distributed evolutionary computation (EC) experiment using cloud resources as
servers, and browsers as clients.
% (Paloma) It seems that you've started the introduction twice. The
% first two paragraphs "introduce the introduction", and so does the
% third. It is specially noticeable when you find "evolutionary
% computation (EC)" twice.
% Yep, I want to say it all, but don't know how to put it together.

%PABLO: He añadido el siguiente párrafo para el issue #31
There exist other volunteer computing frameworks written in compiled languages,
such as C++ or Java, that, in principle, can achieve better performance than
JavaScript. However, one of the advantages of {\sf NodIO} architecture is that
clients do not need to install anything on their machines, they just
%Pablo: According English Stack Exchange, you install "ON" a machine (but "IN" a directory)
need to open a web page  
in a compatible, that is, almost any, browser, such as Chrome, Safari or  Firefox, which are
available in almost every desktop or mobile device. Another advantage is
that the source code can be directly examined by volunteers, because it is not
compiled as other options such as Java Applets, ActionScript, or
ActiveX are. Also, as we will discuss in this paper, it can provide some extra
features such as enhanced security or operating system
independence. This system, however, is not intended to substitute
already existing frameworks running on dedicated hardware or the cloud, but to
%Pablo: same here, I learnt this with my paper of EAs "on clusters"
complement them.

%PABLO: hay que enlazar con el siguiente párrafo
%Desktop clients have the same issues as any 
%volunteer computing system, including our % (Paloma) their?
% own: these issues are
%related to security as well as
%safety, which introduce

%Mario: Rewrote: 
%One of the aspects that must be taken into account in volunteer based
%frameworks is its {\em social} aspect in the design, with 
In general, several volunteers will be needed to
overcome the overhead involved in using a client-server architecture
and this number of volunteers is obviously not guaranteed. Therefore, the {\em social} aspect must be taken into account when designing 
volunteer based frameworks, with 
issues related to security, trust, and privacy among others. 
The computing system becomes then a {\em techno-social system}
\cite{vespignani2009predicting}, in the sense that its social aspects
cannot be separated from its technical specifications, with one
feeding into the other. One of these aspects is safety for the user
running it, as well as for the scientist: the user must be reasonably
sure that the 
experiment is not misusing their resources and the scientist must have a
certain safety in the sense that their server is not going to be
attacked via cheating or denial of service attacks.

We have decided to approach security in a pragmatic way. Including
many security features % (Paloma) I'm kind of lost here... Sentences
                       % seem to be unconnected and unrelated...  
% Separated paragraph and substituting wildcard words - JJ
would decrease performance; on the other hand, we publish all data
resulting from the experiment once private information, such as IPs,
has been processed to an untraceable hash. The experiments themselves
are a proof that there are enough users trusting the experiment to
give it their CPU time and that there has been no attempt to cheat by
sending specially crafted chromosomes or to attack the server to make
it stop working because there was no incentive in doing so.
% (Paloma) Is this really true? Can we really state that there were no
% attacks because there was no incentive? Attackers don't always need
% an incentive to perform an attack, or the incentive could not be
% perceived by us. And why do the experiments prove this? Are there
% specific measures? 
% There were no attacks, that's a fact. There could be in the future,
% but not for the time being - JJ
The
system, besides, is running in a dedicated Platform as a Service (PaaS)
container or Virtual Private Server so accessing the server would only
show the few data obtained from the experiment, adding security against attackers. In addition, making it
free software ensured the meager bounty obtained by attackers were
known by anyone participating in the experiment, via 
links in the web page that hosted it. % (Paloma) What was known? The
                                % fact that it was running in a PaaS?
                                % Why was that known by links?
%That there wasn't anything to hunt. Cleared, I expect.

In this paper, we will carry out three experiments, after establishing
the baseline of the performance that can be expected in a JavaScript
based system and its comparison with other languages in
\cite{2016arXiv160101607Manom}. First, several experiments are
performed using a simple Ajax client that does not 
restart when a solution is found. Second, we change the client a bit
so that it uses web workers, that is, it runs in the background
process and then it restarts when the solution is found. Third, we use the same
% Pablo: it restarts when the solution is found (it has not been
% mentioned yet)
% Just do it! - JJ
client, but with a harder problem that cannot be easily solved by a single
volunteer and thus needs the contribution of many volunteers to
proceed. 
%Mario: refrased: 
%With these three experiments we want mainly to try and model
With these three experiments our main objective is to try and model
the behavior of users and learn how different problems and
implementation choices have an influence on it. That is why in this
paper we will collect information on the number of users participating
in each experiment and also the time they spent doing so. In this
paper we will show how this participation follows certain patterns,
and that they depend on the implementation itself as well as the
problem and its difficulty; however, certain implementations allow to
extract more CPU cycles from volunteers.

%Pablo: To do this we compare the time spent by each user because...
% Pablo, please, don't JJ me. Do the changes, because I'm not learning
% now. I have added a bit anyway.

@@

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX, or Java applets were
used with the same purpose; for instance, Java was pointed out in \cite{soares1998get} as a
``language for the internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms, and was proposed in \cite{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using compiled Java
programs. The ATLAS system
\cite{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. 
%% We need to only reference this section?:
%%------------------- BEGIN
% This paper % (Paloma) No really need for that "also", plus it complicates reading
% establishes the desirable features of
% any {\em global computing} infrastructure, that can also be applied to
% distributed evolutionary computation experiments that rely on
% volunteer computing: scalability, heterogeneity, fault tolerance,
% adaptive parallelism, volunteer's safety, anonymity, hierarchy, ease of use, and
% reasonable performance. But hierarchy, this is, the capability of
% preferentially using resources in the same organization, is a feature
% that could be either disregarded or included within adaptivity, which
% leaves us with eight desirable properties. 
% % (Paloma) Re-write. For example, "But is hierarchy, this is, the
% % capability of preferentially using resources in the same
% % organization, the one that could be either disregarded or included
% % in adaptivity, which leaves us with eight desirable properties." 
% % (Paloma) And I re-wrote what I proposed yesterday.
% % thanks :-) -JJ
% %%------------------- END
% %% EDIT:
% %%------------------- BEGIN
% The fulfillment of these properties could be achieved by using any of the
% languages and associated technologies mentioned above, provided that the user had
% installed the required virtual machine plugin first. 
% %%------------------- END
The fulfillment of the desirable features of any distributed evolutionary 
computation experiments that rely on volunteer computing \cite{2016arXiv160101607Manom} could be achieved by using any of the
languages and associated technologies mentioned above, provided that the user had
installed the required virtual machine plugin first.

% (Paloma) These should be separate sentences
However, embedding a
distributed computation client in the browser has the advantage of
providing a single and universal way of accessing resources via an
URI, something that is not guaranteed by using desktop applications. 
Baratloo et al. tried in 1996 to create a framework with
the aforementioned qualities called {\em Charlotte}
\cite{baratloo1999charlotte}. This framework, which implemented a
distributed shared memory, was inspired by a previous work
on parallel computing libraries PVM and MPI and solved most of the issues by
using the Java virtual machine embedded in the browser,
which provided
security to the user, uniformity of computing environment
independently of the operating system, and access to a common data store in the
server. Additionally, since the Java virtual machine is a sandbox, it was safe to the user
and as it runs compiled bytecode it also guaranteed a certain level of
performance. % (Paloma) I've changed a bit this sentence to unify the use of past tense when talking about Charlotte
Other issues such as scheduling, which are related to adaptive 
parallelism, were solved on the server. But it is interesting to note that,
even in the early stages of the web, the possibility of
% (Paloma) Re-write. For example, "but it is interesting to note that, even in the early stages of the web, the possibility of..."
browser-based distributed computing was already researched.

Other systems, like JET, proposed by Soares et
al. \cite{soares1998get}, also used Java. JET supports
the execution of parallel applications over the Internet and features
a comprehensive statistics collection subsystem, enabling its use for science, %(Mario) Was easier to change "allowing"
since it allows % (Paloma) Use synonyms, like "yield" or "permit" 
the collection of algorithm analytics by the
experiment designer, unlike the other systems mentioned
above. Other proposals contemporary to JET, such as {\em SuperWeb}
\cite{alexandrov1997superweb} are roughly a translation of classical
distribution techniques, with overly complicated architectures that include
brokers for resource discovery. SuperWeb is focused,
% (Paloma) Careful with starting using plural ("Other proposals", "are") and continuig with singular. If you are referring to "SuperWeb", change "It" for "SuperWeb"
however, in solving the
issues involved in browser-based computing and also in making a system
that is able to work across different types of computers. Interestingly enough, this
paper incorporates a discussion on the economic model, paving the way
for the for-profit distributed computing systems that arose later on. 
On the other hand, the potential for volunteer computing using
browsers was realized
later on \cite{sarmenta-bayanihan} as well as the potential of
user misconduct when having access to the source code \cite{sarmenta-sabotagetolerance}. 
However, these early efforts by
Sarmenta once again used Java and not JavaScript, making them % (Paloma) making them
less widespread.

The main problem with using Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense that an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser.
% Also these virtual environments have become targets of malicious attacks; 
% for instance numerous vulnerabilities have been discovered in 
% the proprietary Adobe Flash runtime environment and
% video player \cite{ford2009analyzing,watanabe2010new}. 
The need for
these additional environments has been mitigated with the
adoption of the HTML5 standard \cite{anthes2012html5}, which is
tightly integrated with JavaScript.

JavaScript can be used either for unwitting, that is, stealth,
\cite{unwitting-ec} or volunteer 
\cite{langdon:2005:metas,gecco07:workshop:dcor} distributed
evolutionary computation and it has been used ever since by several
authors, including more recent efforts
\cite{duda2013distributed,DBLP:journals/corr/abs-0801-1210}.

% that even
% used the client's GPU \cite{duda2013gpu}. 
%Also a bit out of scope. Have to reduce the paper. 
% Even if, as said above, Java
% is not the best option, many other researchers have
% used Java \cite{chong:1999:jDGPi} and others have gone away from the
% server-based paradigm to embrace peer to peer systems
% \cite{jin2006constructing,10.1109/ICICSE.2008.99,DBLP:conf/3pgcic/GuervosMFEL12}. These
% computing 
% platforms avoid single points of failure (the server) but, since they
% need a certain amount of infrastructure installed to start, the
% threshold to join them is much higher. % (Paloma) Do yo
                                % mean... higher? A low threshold
                                % means that is easy to happen
                                % something.
%Correct. But maybe in this new version we should eliminate this
%paragraph altogether. 
% The next paragraph has been suppresed.

%Also suppressing this one which is probably not needed in this paper
%- JJ
% JavaScript is also compatible with the functional programming paradigm \cite{Cousineau1998,MacLennan1990,Thompson1996}.
% According to \cite{swanresearch2015}, the benefit of using this kind of paradigm in
% metaheuristics optimization is that it allows an improvement in communicability,
% reproductibility, interoperability, automated assembly, knowledge engineering,
% and efficiency. In the work presented in this paper, JavaScript has been used throughout the full
% stack, which has the advantage of using the self-same lines of code for
% writing an evolutionary algorithm for the
% desktop or command line and browser clients, both based in the EC
% library {\sf NodEO} \cite{DBLP:conf/gecco/GuervosVGES14}. This saves time 
% when writing code and leverages 
% the expertise acquired
% % (Paloma) Not sure if "acquired" should be written before
% % "expertise"... sounds better to me, anyways
% % Don't know what you mean. Write it down and we'll see - JJ
%  in the library itself.

Additionally, we are also interested in measuring the performance
of volunteer computing systems, an area in which there have been
relatively few efforts. There were some initial attempts to avoid the differences in performance
that could be obtained from volunteers  by making
the algorithm adaptive to the kind of resources allotted to it
\cite{milani2004online}. This is actually not such a big problem in
algorithms such as the Evolutionary Algorithm (EA) that can easily be
parallelized via population splitting or by farming out evaluations to all
available nodes. Lately, several approaches have focused on the
fault-tolerance of volunteer algorithms
\cite{gonzalez2010characterizing} which can, of course, be studied in
the more general context of distributed computing 
\cite{nogueras2015studying} or be included in a broader study of the
performance of the EA itself
\cite{DBLP:journals/gpem/LaredoBGVAGF14}.

Moreover, measuring the performance of the resulting metacomputer
involves understanding its dynamics. Initial
work was done for peer to peer systems by Stutzbach et
al. \cite{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \cite{churn08,laredo2008rcp}. 
%But the raw material of           
%volunteer computing, % (Paloma) Whith regard to?  Maybe it was colloquial/analogy
But some of the essential metrics in volunteer computing like the
number of users or the time spent by every one in the
computation in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\cite{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\cite{javadi2009mining} found out that the Weibull, log-normal, and
Gamma distributions 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \cite{agajaj}. % (Paloma) I have a doubt here, should this reference be anonymised or not?
Although SETI@home works
at a larger scale in both quantity of volunteers and the scope of projects, 
our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. 
%Mario: erase: Next we describe the framework itself.
As far as we know, this paper presents one of the few experimental
results using a volunteer computing system which uses  % (Paloma) Change to "which" in order to not repeating "that" all the time
very diverse computational resources; these will be described next.

\section{Methodology and experimental results}
\label{sec:exp1}

% A bit of description of the framework, with reference to published
% reports - JJ

We set up different experiments by deploying the server in free cloud
services, changing it from one experiment to the next if needed. This server runs a web service
which stores chromosomes sent from clients and serves requests to
them; it is 
an open source system that has been extensively described in
\cite{2016arXiv160101607Manom}. The client runs an evolutionary
algorithm written in JavaScript that after a number of generations
sends the best individual back to the server, requesting at the same
time a random individual of those stored in the server. The experiment
starts by setting up a website that is sent by the server and then announcing it on several social
networks so that people contribute to it by visiting the page and
staying there for a while.

\subsection{Initial tests with a simple Ajax client}
%% Above we mention three experiments:  Simple Ajax client, Web Workers, Harder Problem
%% Im renaming subsection to  reference the first experiment

Initial tests with an evolutionary algorithm solving the 40-Trap
problem were set up using the free PaaS OpenShift; the results are described in
Table \ref{tab:summary:os}. In general, every test took several
days and during its performance the problem was solved more than 50 times with
several hundreds different IPs participating in it; each
evolutionary algorithm stopped when the solution was found. 
%
\begin{table*}
\caption{Summary of time per run, number of IPs and number of PUTs per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccccc}
\hline
Date & Experiments &\#IPs &\#IPs &  \#IPs & Time (s) &  \#PUTs & $<$ 69s & $<$ 3.46s & Inter-experiment\\
     & \# & Different & Median & Max & Median &  Median & & & correlation \\ 
\hline
4/4 & 57 & 191 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 & 231 & 559 & 5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 97 & 179 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%

The second-to-last column in Table \ref{tab:summary:os} shows the percentage of experiments in which running the
experiment in the browser takes {\em less} time than in the
desktop. Volunteer computing systems do not exclude desktop
computing and, in fact, the server can handle any client, not only
those running JavaScript on the browser. Although, we include it here for the
sake of comparison and to show that, being a free resource, it can
anyway make a valid and valuable contribution to any distributed or
desktop computing experiment. Another interesting result is that the
correlation between users of an experiment and the next is
non-existent, that is, computing nodes are not {\em
  staying} after the solution has been found. This is the issue that we
address in the next experiments. 
%Mario: changed: will address in this paper.

% \begin{figure}[!htb]
% \centering
% \includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
% \includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
% \includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
% \caption{Duration of experiments vs. number of different IPs (nodes)
%   participating in it, with averages and standard deviation shown as
%   black dots and lines, single dot if it was 1). Shade of gray or
%   color indicate how many experiments included that many unique IPs.
% From left to right, experiments 4/4, 4/24 and 7/31.}
% \label{fig:duration}
% \end{figure}
% % ../data/time-vs-IPs-openshift.R
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color (or gray value) is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
% computed via ../data/time-vs-IPs-openshift.R
\end{figure}
%
% We will have to analyze experimental data a bit further to find out why
% this happens and also if there are some patterns in the three sets of
% experiments. An interesting question to ask, for instance, is if
% by adding more computers makes the experiment take less. In fact, as
% shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
% not seem to contribute to decreasing the time needed to finish the
% experiment. However, the cause-effect relationship is not clear at
% all. It might be the opposite: since experiments take longer to finish
% and might in fact be abandoned with no one contributing for some time,
% that increases the probability of someone new joining them. In fact,
% with experiments taking a few seconds and due to the way the
% experiments are announced, it is quite difficult that several
% volunteers join in in such a short period of time, even more if we take
% into account that volunteers are not {\em carried over} from previous
% experiments. This implies that it would be convenient to use a problem
% of a bigger size to check this hypothesis as we have done in the next
% experiment. 
%  However, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{puts-openshift-4-4.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-4-24.png}
\includegraphics[width=0.32\linewidth]{puts-openshift-7-31.png}
\caption{Number of PUTs per unique IP and fit to a Generalized
  Extreme Value distribution (in a lighter shade or blue). From left to right, experiments
  4/4, 4/24 and 7/31.} 
% Plotted with ../data/plot-zipf-openshift.R
\label{fig:puts:os}
\end{figure}
%
It is also interesting to check the distribution of the experiment
duration, shown in Figure \ref{fig:zipf:os} and which roughly follows
a Zipf's law, with similar distribution along all three runs. The 4/24
run is the most complete and shows an S-shape, which implies an
accumulation of experiments taking similar time and around 100
seconds. The most interesting part is the {\em tail}, which shows how
many experiments took a desirable amount of time, on the order of
10 seconds, and which appears in all three graphs. As it can be seen,
it sharply drops implying there are 
just a few of them, and with diminishing probability as time
decreases. This exponential distribution also appears in Figure \ref{fig:puts:os}, which shows the
distribution of HTTP PUTs, equivalent to the number of
generations divided by 100, contributed by every user. These results show a Zipf-like behavior,
so we have fitted it to the Generalized Extreme Value distribution,
with the resulting parameters shown in Table \ref{tab:puts:os}.
%
\begin{table}
\caption{Summary of fit to Generalized Extreme Value distribution of
  the number of PUTs per unique IP. \label{tab:puts:os}}
\begin{center}
\begin{tabular}{l|ccc}
\hline
Date  & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
4/4 &  8.541 $\pm$ 1.0926  &    12.442 $\pm$ 1.7302 &  1.388 $\pm$
0.1377 \\
4/24 & 6.148 $\pm$ 0.3782 & 7.354 $\pm$ 0.5105 & 1.090 $\pm$  0.0697  \\
7/31 & 11.645 $\pm$ 1.475 & 16.365 $\pm$ 2.201 &  1.265 $\pm$ 0.132   \\
\hline
\end{tabular}
\end{center}
\end{table}
%
This distribution was originally proposed to fit extreme values
\cite{resnick2013extreme} and contains, as a special case, the inverse
Weibull distribution which was fitted to volunteer computing
frameworks such as SETI@home \cite{javadi2009mining}. We obviously do
not pretend to compare our framework in scale or complexity with it, but
to point out that the behavior of volunteer computing nodes follows a
certain pattern, found in SETI@home, and which also appears in our framework. This
distribution is governed by three parameters, the usual location $\mu$
which is related to where it has its {\em center} and a scale $\sigma$,
related to the size, but also a third shape $\xi$ parameter that is
related to its skewness, that is, how skewed it is around the central
location. Positive parameters indicate that the distribution {\em
  leans} towards the origin, and negative ones towards the other extreme
value. In this case, Table \ref{tab:puts:os} shows $\xi$ values
greater than one and between one and 1.4, which indicates that the
three experiments share this origin-leaning pattern, with many users
donating a few cycles and just a few donating extreme values. Random
distributions with these parameters have been plotted in red in
Figure \ref{fig:puts:os}, indicating that the fit is good enough. The
predictive value of these fits, however, is limited, over all taking
%Pablo: question for Paloma: is over all or overall?
into account the low correlation between successive events shown
above.

At any rate, this also shows that a convenient way of increasing the
%Mario: As I understend is: 
computing power would be increasing this minimum amount of experiments
per user. For this reason, we made some changes to the implementation,
whose results are presented in the next subsection. % (Paloma) This sentences are sort of alone here... Could we extend a bit the first "this" to connect them to the previous paragraph ?

\begin{table*}[!htb]
\caption{Summary of {\sf NodIO-W$^2$} (an acronym for the version that
  uses web workers) and comparison with the previous
  experiments, which have been aggregated to compute central measures. \label{tab:summary:ww}}
\begin{center}
\begin{tabular}{l|ccccccc}
\hline
Date & Median \#IPs & Max \#IPs & Median time (s) & Median \# PUTs & $<$ 69s & $<$ 3.46s & Inter-experiment correlation\\
\hline
{\sf NodIO} & 5 & 29 & 123 & 14 & 37.43\% & 3.40\% & 0.10 \\
{\sf NodIO-W$^2$} & 4  & 16 & 7.36 & 40 & 89\% & 36.90\% & 0.4336061 \\
\hline
\end{tabular}
\end{center}
\end{table*}
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.9\linewidth]{ips-time-ww.png}
\caption{Time employed in every experiment vs. number of IPs (in
  abscissas). The black dot and line show the average and standard
  deviation, the blue to red shades the number of experiments with the same
  number of IPs. } 
\label{fig:ipstime:w2}
% ../data/time-vs-IPs.R 
\end{figure}

\subsection{Experiments using web workers}
\label{sec:w2}

The results reported above indicated that we should look for a way of
increasing the time users spend on the page. This led us to create an
architecture that uses web workers, which is described in
\cite{2016arXiv160101607Manom} and we abbreviate to {\sf
  NodIO-W$^2$}; web workers work in the background even if
the web page is not in focus. A few improvements were also tested,
basically using a random population and restarting the web worker when
the solution was found.

The experiments were run in the same way as before, using several
 announcements in Twitter throughout several days by the end of July
2015. Eventually, more than one thousand experiments were completed in
a matter of days. A summary of results is shown in Table
\ref{tab:summary:ww}, comparing it with the aggregate of the results
obtained with the initial version of {\sf NodIO}. These results are
remarkably different, being similar only in the median and maximum
number of IPs, although in this case it would be combination
IP-worker, as we considered every worker as a unique IP. The median
number of seconds is almost 80\% better than the previous version, and
the median number of PUTs is three times better; this causes that most
of experiments take less time than one of the baselines, and one third
less than the second baseline. Thus, the conclusion in this case is
that we can obtain, through volunteer computation, a result that most 
of the times is better than we would by using a similar, non-parallel,
desktop setup, which makes this system suitable for massively
distributed evolutionary algorithms.
%
\begin{table}
\caption{Summary of fit to GEV and Weibull distribution of
  the number of PUTs per worker. \label{tab:puts:ww}}
\begin{center}
\begin{tabular}{cccc}
\hline
Distribution & Location $\mu$ & Scale $\sigma$ & Shape $\xi$ \\
\hline
GEV & 18.275 $\pm$ 0.837  &  25.96  $\pm$ 1.22 & 1.242   $\pm$ 0.047 \\
Weibull & ND & 75.11 $\pm$ 2.97  & 0.697 $\pm$ 0.014 \\
\hline
\end{tabular}
\end{center}
\end{table}
%

However, it is interesting to note why this is so, and the first hint
is the inter-experiment correlation between the number of IPs in
successive experiments. While before it was an unremarkable 10\%, it is
now more than 43\%, making the number of IPs in contiguous experiments
highly correlated. The main reason for this is the setup in the
version using web workers, which restarts an island after a solution has been
found and also that it keeps running even if the tab is not on the
foreground. This means that volunteers can leave the experiment
running for as long as they want and they will be contributing 
experiment after experiment, unlike the previous version, where they
stopped contributing after one solution was found. This is also
reflected in the number of PUTs, more than 50\% of the volunteers run the experiment for more than
4000 generations. All things considered, this means than there are
%Mario: many more more threads {\em simultaneously} running?
more threads and many more computers {\em simultaneously} running,
leading to this almost 5-fold increase in running time.

We are also interested in measuring the scaling properties of this
model, that is, the relationship between the number of IPs
participating in an experiment and the time it takes to complete
it. This is shown in Figure \ref{fig:ipstime:w2}, which displays every
experiment in terms of time (in milliseconds) vs. % (Paloma) Isn't "vs." informal?
the number of workers participating
in it. Although there is not a clear trend, the graph seems to say
that the time needed for finding the solution in this evolutionary
algorithm does not depend on the number of workers participating in
it. This does not mean that it is independent on the number of {\em
  simultaneous} users, which is probably the case although it is more
difficult to measure. There seems to be a trend towards a decreasing
standard deviation in the time, with more volunteers adding both {\em
  robustness} to an experiment, and certainty in the time it is taking. % (Paloma) Should "certainty" go in italics? Otherwise it's confusing. 
 However, more experiments would be needed to check this hypothesis.

>>>>

\title{NodIO: a framework for pool-based evolutionary computation}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{5} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
\alignauthor
Juan-J.~Merelo, Pedro A. Castillo,Pablo Garc\'ia-S\'anchez,Paloma de las Cuevas\\
\affaddr{Dept. of Computer Architecture and Technology
  and CITIC, University of Granada, Granada, Spain} \\
\email{(jmerelo|pacv|pgarcia|palomacd)@ugr.es}
\alignauthor
Mario Garc\'ia-Valdez,\affaddr{ Dept. of Graduate Studies, Instituto
Tecnológico de Tijuana, Tijuana, M\'exico}\\
}

@@

@@

@@

\section{The NodIO framework and API}
\label{sec:exp1}
